Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-10 analysis2_manuscript.tex
intro

% bibliography
\usepackage[
    backend=biber,
    natbib=true,
    style=authoryear,
    ]{biblatex}
\addbibresource{references.bib}

@@

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \citep[e.g.,][]{gillespie1991causes}.
    For instance, changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \citep{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn, selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus, it is imperative to jointly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However, this is a challenging task from a modeling perspective \citep{johri2022prospect}. % MFR this is a bit unclear

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.  \kelcomment{maybe also cite https://doi.org/10.1093/bioinformatics/btn522}
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few},\kelcomment{also cite: PMID: 32487519 } \drscomment{and maybe a few others? PMIDs: 24102784 and thornton2019polygenic} to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting},
    to explore the power and limitations of statistical methods \citep[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{beaumont2002approximate,pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}. \kelcomment{also cite: PMID: 36617238 }
    While this is so, joint simulation of complex demography and selection
    is challenging,  requires a deep understanding of the underlying
    evolutionary processes, and necessitates a number of parameter choices including
    the strength of selection and the
    recombination rate, as well as a parameterized demographic model.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, different simulation and inference tools may use different structures
    and parameter scalings for their models,
    making it difficult to directly compare across these tools and the studies that use them.

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation\citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.

>>>>

% bibliography
\usepackage[
    backend=biber,
    natbib=true,
    style=authoryear,
    uniquename=false,
    ]{biblatex}
\addbibresource{references.bib}

@@

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \citep[e.g.,][]{gillespie1991causes}.
    For instance, changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \citep{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn, selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus, it is imperative to jointly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep[e.g.,][]{sheehan2016deep,johri2020toward},
    yet standardized tools for doing so remain limited.

    % simulation
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular, hernandez2008flexible, thornton2014c++, haller2019slim}. 
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{peischl2013accumulation, thornton2019polygenic, galloway2020few, zhang2020impact}, 
     to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting},
    to explore the power and limitations of statistical methods \citep[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods 
    \citep[e.g.,][]{beaumont2002approximate,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic, zhang2023maladapt, smith2023dispersenn2}.
    While this is so, joint simulation of complex demography and selection
    is challenging,  requires a deep understanding of the underlying
    evolutionary processes, and necessitates a number of parameter choices including
    the strength of selection and the
    recombination rate, as well as a parameterized demographic model.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, different simulation and inference tools may use different structures
    and parameter scalings for their models,
    making it difficult to directly compare across these tools and the studies that use them.

    % stdpopsim
    In light of the complexities associated with estimating and simulating
    population genetic models, it is no surprise that
    a lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community 
    \citep[e.g.,][]{speidel2021inferring, wang2021automatic, yang2022trustgwas, degiorgio2022spatially, browning2023fast, schweiger2023ultrafast, temple2024modeling, haag2025pandora}.
    While that is so, previous releases of \stdpopsim did not include
    models of selection, which is a major limitation for more empirically motivated
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation\citep[e.g.][]{thornton2019polygenic,hartfield2024polygenic}, the effects of selective sweeps 
    \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, 
    and the impact of background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Furthermore the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Moreover, one should be able
    to model all of these features simultaneously using estimates and annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background,rodrigues2024shared}.
